Synthesizing Symmetric Lenses

### 2018-06-07 papers/SymmetricLenses/main.tex
react to comments

Unfortunately, writing such synchronizers is tedious and error
prone.\saz{``tedious and error prone'' is such a PL clich\'{e}, can we do better?}
Bidirectional languages ease the difficulty of writing such synchronizers, because a
single, well-typed program expresses a \emph{lens}---a group of conversion
functions guaranteed to satisfy round-tripping laws. However, bidirectional
languages are themselves difficult to program in, often requiring users to permute
data in fiddly ways while satisfying the type system's complex unambiguity
constraints.

@@

\emph{Symmetric lenses} \saz{citations?} formalize this richer class of
non-bijective conversion functions by using additional state, called the
\emph{complement}, to help with synchronization. While increasing
expressiveness, the existence of the complement makes symmetric lenses difficult
to use in real-world synchronization tasks, because users\saz{lens developers?
  lens programs?}  must maintain the synchronized files as well as the
complements. The complement further complicates synthesis, because input-output
examples no longer suffice; instead edit sequences must be provided.

In this paper, we reformulate \saz{this sounds like they may be of the same
  expressiveness, but perhaps we mean to ``constrain'' symmetric lenses?}
symmetric lenses as \emph{simple symmetric lenses}. This reformulation does not
use a complement, and only uses the files that we aim to synchronize as
state. We identify a property, \emph{forgetfulness}, that identifies the subset
of symmetric lenses that can be written as simple symmetric lenses. We provide
combinators for building simple symmetric lenses on string data, and integrate
these combinators with Boomerang, a language for bidirectional programming on
strings. In Boomerang, and our extension to it, lens expressions are typed
between regular expressions, where the regular expressions describe the
synchronized data formats.

@@

\saz{This paragraph is hard to follow --- maybe introduce $\ell : S
  \Leftrightarrow T$ and use the names $S$ and $T$?}
A primary difficulty is that there are a relatively many well-typed simple
symmetric lenses. As symmetric lenses permit data to be present in only one
format, for any data component there is now a choice: is this data only relevant
to this format and as such should be left unaligned (changes to this data are
not propagated to the other format), or is this data present in both formats and
as such should be aligned (changes to this data change are propagated to the
other format).

@@

We evaluate our algorithm on a benchmark suite consisting of synchronizing file
formats from configuration files, application-specific data storage files, and
data cleaning tasks. We find that our algorithm is able to synthesize all
\saz{or each?}  of these lenses in under 20 seconds.

@@

\saz{Perhaps we should explicitly say here that we are trying to define a lens,
  or describing the functions that a lens provides.}
To synchronize these files, a few functions are needed. Two \emph{creates}
\saz{we haven't introduced ``create'' as a noun yet.} are
needed; if a crontab is initialized, a synchronized Scheduled Tasks file should
also be created, and vice versa. Furthermore, \emph{puts} are needed; if a
crontab is updated, we should be able to combine it with a non-updated Scheduled
Tasks file, and output a new Scheduled Tasks file, synchronized to the updated
crontab. To have a simple symmetric lens is to have these required
synchronization functions. Formally, a simple symmetric lens between $X$ and $Y$
comprises 4 functions:
\begin{enumerate}
\item $\CreateR \OfType X \rightarrow Y$
\item $\CreateL \OfType Y \rightarrow X$
\item $\PutR \OfType X \rightarrow Y \rightarrow Y$
\item $\PutL \OfType Y \rightarrow X \rightarrow X$
\end{enumerate}
Furthermore, if a file is not edited, but is then resynchronized using a put
function, no data should be changed. To be a simple symmetric lens, one must
satisfy the following ``round-tripping'' laws, guaranteeing identity edits won't
change the file data.
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{gather}

@@

or(ins("<Tasks></Tasks>"),nonempty_task_map) : crontab $\Leftrightarrow$ scheduled_job
\end{lstlisting}
  \caption{A lens, written in our language for building simple symmetric lenses
    on string data, that synchronizes crontab files and Scheduled Task files.
    Note that we can annotate our lenses with types, to confirm the lens
    operates on the languages of the regular expressions.\saz{Say something
      about the possibility of synthesizing this lens?}}
  \label{fig:example_lens}
\end{figure}

@@

How does \RXSearch propose stochastic regular expression pairs? We define
rewrite rules for stochastic regular expressions, and prove those rewrites
preserve language and probability distributions. By applying these rewrites, we
can generate new regular expression pairs, syntactically distinct but
semantically equivalent to the originals.

@@

\saz{The following is hard to follow}
We address this issue by keeping track of the best lens found thus far, and
eventually cutting the proposal-synthesis loop short. Before proposing new
types, the regular expression generator, \RXSearch, evaluates whether it should
continue proposing regular expression pairs, or whether it is content with the
current best. For a given regular expression pair, \RXSearch looks at the number
of rewrite rules performed on the pair, and identifies how many pairs
involving the same number of rewrites may be proposed. If there are a large
number of pairs that can be proposed with that many rewrites, or the current
lens is very well synchronized, \RXSearch stops. In this
way, \RXSearch will continue searching for better lenses, but does not tackle
huge search spaces unless it needs to. In this way, the algorithm can provide
lenses quickly to the users, which users can refine with additional examples.
% end overview

>>>>

Unfortunately, writing such synchronizers is quite difficult; we must write
parsers both formats, specify how the parsed data is converted into the other
format, and specify how to propagate updates from one format to the other.
Bidirectional languages ease the difficulty of writing such synchronizers,
because a single, well-typed program expresses a \emph{lens}---a group of
conversion functions guaranteed to satisfy round-tripping laws. However,
bidirectional languages are themselves difficult to program in, often requiring
users to permute data in fiddly ways while satisfying the type system's complex
unambiguity constraints.

@@

\emph{Symmetric lenses}~\cite{symmetric-lenses} formalize this richer class of
non-bijective conversion functions by using additional state, called the
\emph{complement}, to help with synchronization. While increasing
expressiveness, the existence of the complement makes symmetric lenses difficult
to use in real-world synchronization tasks, because lens programs must maintain
the synchronized files as well as the complements. The complement further
complicates synthesis, because input-output examples no longer suffice; instead
edit sequences must be provided.

In this paper, we contrain symmetric lenses to \emph{simple symmetric lenses}.
These constrained lenses do not use a complement, and only use the files that we
aim to synchronize as state. We identify a property, \emph{forgetfulness}, that
identifies the subset of symmetric lenses that can be written as simple
symmetric lenses. We provide combinators for building simple symmetric lenses on
string data, and integrate these combinators with Boomerang, a language for
bidirectional programming on strings. Both Boomerang and our extension type lens
terms between pairs of regular expressions. The typing judgment $\Lens : \Regex
\Leftrightarrow \RegexAlt$ means the lens term $\Lens$ encodes synchronizing
functions between the format described by $\Regex$ and the format described by
$\RegexAlt$.

@@

\saz{This paragraph is hard to follow --- maybe introduce $\ell : S
  \Leftrightarrow T$ and use the names $S$ and $T$?} A primary difficulty is
that there are a relatively many well-typed simple symmetric lenses. If we are
trying to find a lens between $\Regex$ and $\RegexAlt$, there is now a choice:
do $\Regex$ and $\RegexAlt$ represent the same data, and should be aligned
(changes to data in the language of $\Regex$ are propagated to synchronized
strings in the language of $\RegexAlt$), or should it be left unaligned (changes
to data in $\Regex$ update synchronized strings in $\RegexAlt$). Or instead
should certain components of $\Regex$ be aligned with certain components of
$\RegexAlt$, with the other components left unaligned.

@@

We evaluate our algorithm on a benchmark suite consisting of synchronizing file
formats from configuration files, application-specific data storage files, and
data cleaning tasks. We find that our algorithm is able to synthesize each of
these lenses in under 20 seconds.

@@

Lenses synchronize files by providing 4 functions, two \emph{create} functions,
and two \emph{put} functions. Create functions are used to initialize data in
one format when provided with data in the other. If a crontab is initialized, a
synchronized Scheduled Tasks file should also be created, and vice versa. Put
functions are used to propagate edits from one format to the other. If a crontab
is updated, we should be able to combine it with a non-updated Scheduled Tasks
file, and output a new Scheduled Tasks file, synchronized to the updated
crontab. To have a simple symmetric lens is to have these required
synchronization functions. Formally, a simple symmetric lens between $X$ and $Y$
comprises 4 functions:
\begin{enumerate}
\item $\CreateR \OfType X \rightarrow Y$
\item $\CreateL \OfType Y \rightarrow X$
\item $\PutR \OfType X \rightarrow Y \rightarrow Y$
\item $\PutL \OfType Y \rightarrow X \rightarrow X$
\end{enumerate}
Furthermore, if a file is not edited, but is then resynchronized using a put
function, no data should be changed. To be a simple symmetric lens, one must
satisfy the following ``round-tripping'' laws, guaranteeing identity edits won't
change the file data.
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{gather}

@@

or(ins("<Tasks></Tasks>"),nonempty_task_map) : crontab $\Leftrightarrow$ scheduled_job
\end{lstlisting}
  \caption{A lens, written in our language for building simple symmetric lenses
    on string data, that synchronizes crontab files and Scheduled Task files.
    Note that we can annotate our lenses with types, to confirm the lens
    operates on the languages of the regular expressions. In this work, we
    synthesize the lens itself from such regular expressions. \saz{Say something
      about the possibility of synthesizing this lens?}}
  \label{fig:example_lens}
\end{figure}

@@

How does \RXSearch propose stochastic regular expression pairs? We define
rewrite rules for stochastic regular expressions, and prove those rewrites
preserve language and probability distributions. By applying these rewrites, we
can generate new regular expression pairs, syntactically distinct but
semantically equivalent to the originals. \RXSearch proposes the regular
expression pairs which are most similar to the originals first (in that they
require fewer rewrite rules to generate), and iteratively propose pairs that are
increasingly dissimilar to the original pair (in that they require more rewrite
rules to generate).

@@

\saz{The following is hard to follow} Eventually, we cut the proposal-synthesis
loop short. Before proposing new types \RXSearch evaluates whether it should
continue proposing regular expression pairs, or whether it is content with lens
of lowest cost seen thus far. If there are a large number of regular expression
pairs that are equally dissimilar to the originals or if the current lens is
very well synchronized, \RXSearch stops. By doing this \RXSearch continues
searching for better lenses, but does not tackle huge search spaces unless it
needs to. In this way, the algorithm can provide lenses quickly to the users,
which users can refine with additional examples.
% end overview
